### 2024-08-10 paper.tex
more about centralizer

\documentclass[12pt,psamsfonts]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{blkarray}
\usepackage{upquote}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{url}
\usepackage{tikz-cd}
\usepackage{enumerate}

@@

>>>>

\documentclass[12pt,psamsfonts]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{blkarray}
\usepackage{upquote}
\usepackage{tikz}
\usepackage{bbm}
\usepackage{url}
\usepackage{tikz-cd}
\usepackage{enumerate}
\usepackage{rotating}

\newcommand{\leftshift}{\mathbin{\rotatebox[origin=c]{0}{$\ll$}}}
\newcommand{\downshift}{\mathbin{\rotatebox[origin=c]{90}{$\ll$}}}
\newcommand{\rightshift}{\mathbin{\rotatebox[origin=c]{180}{$\ll$}}}
\newcommand{\upshift}{\mathbin{\rotatebox[origin=c]{270}{$\ll$}}}

@@

\section{Finding a centralizer}
For nilpotent \(X\) in Jordan form and \(a\) in `normalized' form (i.e., with at most one nonzero element, which is a one), we will find the centralizer of \(A_{X, a, b}\) in 
\[\{(A, I) : A \in \gl_m\} \subseteq \gl_{m + n}.\]
Note that an element of the form \((A, I)\) commutes with \(A_{X, a, b}\) if and only if \(A\) commutes with \(X\), and \(aA = a\), and \(Ab = b\).
So, we just have to find 
\[\{A \in \gl_m : AX = XA, aA = a, Ab = b\}.\]
We begin by finding 
\[\{A \in \gl_m : AX = XA\}.\]
We write the shape of \(X\) as \(\lambda = (\lambda_1, ..., \lambda_k)\), so that 
\[X = \begin{pmatrix}
    J_{\lambda_1} & \\
    & \ddots \\
    & & J_{\lambda_k}
\end{pmatrix}.\]
Then we write \(A\) as the block matrix
\[A =: \begin{pmatrix}
    A_{11} & \cdots & A_{1k}\\
    \vdots & & \vdots\\
    A_{k1} & \cdots & A_{kk}
\end{pmatrix},\]
where \(A_{ij}\) is a block of size \(\lambda_i \times \lambda_j\).
We have 
\[XA = \begin{pmatrix}
    J_{\lambda_1}A_{11} & \cdots & J_{\lambda_1}A_{1k}\\
    \vdots & & \vdots\\
    J_{\lambda_k}A_{k1} & \cdots & J_{\lambda_k}A_{kk}
\end{pmatrix}, \textrm{ and } AX = \begin{pmatrix}
    A_{11} J_{\lambda_1} & \cdots & A_{1k} J_{\lambda_k}\\
    \vdots & & \vdots\\
    A_{k1} J_{\lambda_1} & \cdots & A_{kk} J_{\lambda_k}
\end{pmatrix}.\]
So, the constraint that \(XA = AX\) is simply saying that 
\[\forall i,j. \; J_{\lambda_i} A_{ij} = A_{ij} J_{\lambda_j}.\]
It's easy to see that left multiplication by \(J_{\lambda_i}\) is just a down-shift by one, and right multiplication by \(J_{\lambda_j}\) is a left-shift by one.
This means that \(A_{ij}\) is constrained to be a ``lower-left-Toeplitz'' matrix.
A Toeplitz matrix is one which is constant along diagonal bands: \(\forall ijk. \; x_{ij} = x_{i + k,j + k}\).
A lower-left-Toeplitz matrix is one which is all zeroes except for the bottom-left corner; that is, the bands start in the bottom-left corner, and continue until hitting the band which includes the lower-right corner or the band which includes the upper-left corner (whichever comes first).
In other words, an \(n \times m\) lower-left-Toeplitz matrix is one which satisfies \(x_{ij} = 0\) for all \(i,j\) with \(i < j\) and all \(i,j\) with \(i - j < m - n\).
In yet other words, a matrix is lower-left-Toeplitz if it is Toeplitz, satisfies \(x_{1j} = 0\) for \(j \neq 1\), and satisfies \(x_{in} = 0\) for \(i \neq m\).
Anyway, it is easy to see that the lower-left-Toeplitz matrices are exactly those matrices \(A\) such that left-shifting \(A\) by one has the same result as down-shifting by one.
\par So, we have found the set 
\[C_1 := \{A \in \gl_m : AX = XA\}.\]
It is just the set of \(A = [A_{ij}]_{ij}\), where each \(A_{ij}\) is lower-left-Toeplitz.
\par Let \(v_{ij}\) be the leftmost column of \(A_{ij}\), so that
\[A_{ij} = \begin{pmatrix}
    v_{ij} \downshift 0 & \cdots & v_{ij} \downshift [\lambda_j - 1]
\end{pmatrix}.\]
Since \(A_{ij}\) is lower-left-Toeplitz, \(v_{ij}\) is of the form 
\[v_{ij} = \begin{pmatrix}
    0\\
    v_{ij}'
\end{pmatrix},\]
where \(v_{ij}' \in \mathbb{C}^{\min(\lambda_i, \lambda_j)}\) can be chosen freely.
Now we determine which matrices of this form satisfy \(aA = a\).
For simplicity, we will instead find the matrices \(A\) such that \(aA = 0\) (so that \(a(A + I) = a\)).
(Note that \(I \in C_1\), and \(C_1\) is closed under addition, so this is really the same thing.)
\par In the case that \(a = 0\), clearly every \(A\) works.
Otherwise, let \(i_0, j_0\) be such that \(a_{i_0,j_0} = 1\).
Now, clearly the constraint that \(aA = 0\) is just saying that the \((i_0,j_0)\)th row of \(A\) must be zero.
That is, for each \(j\) the \(j_0\)th row of \(A_{i_0j}\) must be zero.
This just requires that for each \(j\), we must have 
\[v_{ij} = \begin{pmatrix}
    0\\
    v_{ij}'
\end{pmatrix},\]
where \(v_{ij}' \in \mathbb{C}^{\min(\lambda_i - j_0, \lambda_j)}\) can be chosen freely.
So, we have now found the set 
\[C_2 := \{A \in \gl_m : AX = XA, aA = a\}.\]
It is the matrices of the form \(I + A\), where \(A_{ij}\) is a block matrix of size \(\lambda_i \times \lambda_j\) with 
\[A_{ij} = T(v_{ij}) = T\begin{pmatrix}
    0\\
    v_{ij}'
\end{pmatrix},\]
where \(v_{ij}' \in \mathbb{C}^{\min(\lambda_i, \lambda_j)}\) in the case \(i \neq i_0\) and \(v_{ij}' \in \mathbb{C}^{\min(\lambda_i - j_0, \lambda_j)}\) if \(i = i_0\).
\par Now we are only left with finding the matrices \(A \in C_2\) satisfying \(Ab = b\).
This just means finding the matrices \(A\), of the form given above, which satisfy \(Ab = 0\).
We may restate this as the condition that for every \(i\),
\[\begin{pmatrix}
    A_{i1} & \cdots & A_{in}
\end{pmatrix}b = 0.\]
Taking the transpose, 
\[b^\top \begin{pmatrix}
    A_{i1}^\top\\
    \vdots\\
    A_{in}^\top
\end{pmatrix} = 0.\]
Now, the rightmost column of \(A_{ij}^\top\) is \(v_{ij}\) upside-down (rows reversed).
Write \(u_{ij}\) to refer to \(v_{ij}\) upside-down, and \(u_{ij}'\) to refer to \(v_{ij}'\) upside-down.
Then the second-rightmost column is \(u_{ij} \upshift 1\), and so on until the first column is \(u_{ij} \upshift [\lambda_i - 1]\).
So, we can reformulate this condition by requiring that for all \(l\),
\[b^\top \begin{pmatrix}
    u_{i1} \upshift l\\
    \vdots\\
    u_{in} \upshift l
\end{pmatrix} = 0.\]
Writing
\[b = \begin{pmatrix}
    b_1\\
    \vdots\\
    b_k
\end{pmatrix},\]
clearly this is equivalent to requiring that for all \(l\),
\[\begin{pmatrix}
    b_1^\top \rightshift l & \cdots & b_k^\top \rightshift l
\end{pmatrix} \begin{pmatrix}
    u_{i1}\\
    \vdots\\
    u_{ik}
\end{pmatrix} = 0.\]
Then writing \(b_j = \begin{pmatrix}
    b_j'\\
    \vdots
\end{pmatrix}\), where \(b_j'\) has the same dimension as \(u_{ij}\), we can rewrite this constraint as 
\[\begin{pmatrix}
    b_1'^\top \rightshift l & \cdots & b_k'^\top \rightshift l
\end{pmatrix} \begin{pmatrix}
    u_{i1}'\\
    \vdots\\
    u_{ik}'
\end{pmatrix} = 0.\]
Finally, requiring the above equation for all \(l\) is the same thing as requiring
\[\begin{pmatrix}
    b_1'^\top \rightshift 0 & \cdots & b_k'^\top \rightshift 0\\
    \vdots &  & \vdots \\
    b_1'^\top \rightshift [\lambda_i - 1] & \cdots & b_k'^\top \rightshift [\lambda_i - 1]
\end{pmatrix} \begin{pmatrix}
    u_{i1}'\\
    \vdots\\
    u_{ik}'
\end{pmatrix} = 0.\]
This is the most that we will simplify the constraint.
Recall that the \(v_{ij}'\) can be chosen arbitrarily, so the \(u_{ij}'\) can be chosen arbitrarily as well.
That is, the set \(C_3 := \{A \in \gl_m : AX = XA, aA = a, Ab = b\}\) is the set of \(I + A\) for any \(A\) that can be recovered from \(u_{ij}'\) meeting the above condition.
\par Now we find the dimension of \(C_3\).
Since the \(u_{ij}'\) can be chosen arbitrarily, apart from the condition above, we just need to add up their dimensions and subtract the rank of the matrix constructed of the \(b\)'s.
